Polynomial Chaos for the Estimation of Manning's Friction

### 2013-12-11 paper/analysis.tex

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
We now use the surrogate created effectively using PC to propagate prior 
uncertainty through the forward model. We mainly exploit the surrogate 
to study the statistics of the water surface elevation, and to conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.
We note that the statistical and sensitivity analysis presented below 
is performed using the PC coefficients solely; these were calculated
from an ensemble of 125 \geoclaw simulations.
\subsubsection{Statistical analysis}
The mean of a QoI and its standard deviation is computed
from the PC coefficients as shown in Equation~\ref{eq:mean} and Equation~\ref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations in Figure~\ref{fig:rlzs}.
We observed that the standard deviation ......

>>>>

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
We now use the surrogate created effectively using PC to propagate prior 
uncertainty through the forward model. We mainly exploit the surrogate 
to study the statistics of the water surface elevation, and to conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.
We note that the statistical and sensitivity analysis presented below 
is performed using the PC coefficients solely; these were calculated
from an ensemble of 125 \geoclaw simulations.
\subsubsection{Statistical analysis}
The mean of a QoI and its standard deviation is computed
from the PC coefficients as shown in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations in Figure~\ref{fig:rlzs}.
We observed that the standard deviation ......

### 2013-12-11 paper/mcmc.tex

The observed data collected at different locations 
in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure \ref{fig:mcmc} and appears to be well mixed.

>>>>

The observed data collected at different locations 
in the likelihood function (Equation~\eqref{eq:likelihood}) to update the input parameters.
MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure~\ref{fig:mcmc} and appears to be well mixed.

### 2013-12-11 paper/results.tex

\section{Results}
\label{sec:results}
The results of the forward and inverse problems are presented and 
discussed in three sections. First, we perform an error analysis 
of the PC expansion construction, namely  to establish the 
validity of the corresponding surrogate.  Second, we perform a 
sensitivity analysis and rank the impact of the different Manning's 
roughness coefficient on the predicted surface elevation. Finally, 
posterior distributions for the input parameters are determined 
and analyzed in light of the available gauge data.
\input{pcstudy}
\input{analysis}
\input{mcmc}

>>>>

\section{Results}
\label{sec:results}
We begin the results section by performing
an error and convergence analysis of the
constructed surrogate using the PC expansions
to establish its validity in Section~\ref{sec:analysis}.
Then in Section~\ref{sec:forward}, we present
statistical analysis to quantify the uncertainty in the predicted surface elevation
as well as a sensitivity analysis to rank the impact of the different Manning's 
roughness coefficients on this uncertainty. 
Finally, in Section~\ref{sec:inverse}, we present the results of the inverse
problem where we determined the posterior distributions of the input parameters 
and analyzed them in light of the available gauge data.
\input{pcstudy}
\input{analysis}
\input{mcmc}

### 2013-12-11 paper/uqpce.tex

The four-dimensional posterior in Equation~\ref{eq:post_coef} can be explored via MCMC;
this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
once for every proposed set of parameters of the Markov chain. While a single \geoclaw simulation
takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
to avoid forward model simulations altogether. This can be accomplished by constructing a 
surrogate model that requires a much
smaller ensemble of \geoclaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
Polynomial Chaos expansions for accelerating Bayesian inference in this context 
by building a surrogate, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.

@@

The series representation Equation~\ref{eq:stochseries} can be viewed as a spectral expansion
of $U$ along the stochastic dimensions. It can be also seen as
combining approximation and probabilistic frameworks, a combination
that has proven extremely useful in solving UQ problems. The existence and convergence of this series
is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is infinitely smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. In practice the series convergence is monitored 
via various error metrics as discussed in the results section.

>>>>

The four-dimensional posterior in Equation~\eqref{eq:post_coef} can be explored via MCMC;
this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
once for every proposed set of parameters of the Markov chain. While a single \geoclaw simulation
takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
to avoid forward model simulations altogether. This can be accomplished by constructing a 
surrogate model that requires a much
smaller ensemble of \geoclaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
Polynomial Chaos expansions for accelerating Bayesian inference in this context 
by building a surrogate, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.

@@

The series representation Equation~\eqref{eq:stochseries} can be viewed as a spectral expansion
of $U$ along the stochastic dimensions. It can be also seen as
combining approximation and probabilistic frameworks, a combination
that has proven extremely useful in solving UQ problems. The existence and convergence of this series
is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is infinitely smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. In practice the series convergence is monitored 
via various error metrics as discussed in the results section.
